---
title: Moving Files Around In R
author: Xiaoke Liu
date: '2019-08-20'
slug: moving-files-around-in-r
categories:
  - R
  - Automation
tags:
  - R
  - Unix
  - Automation
description: ''
thumbnail: ''

output:
  blogdown::html_page:
    toc: true
---

# Intro  

Recently, I have a project requires some automation. By googling and reading answers on StackOverFlow, I learned quite a lot, I think it's better to document these little tricks here. The automation can be decomposed into below operations:

1. Creating 24 folders with similar naming patterns  
2. Copy same code into 24 folders
3. Change each code's contents slightly, like input dataset name, output dataset file location, etc. 
4. Run same UNIX command on different directory 
5. Covert each code's output files from xls to xlsx 

# Creating 24 folders with similar naming patterns 

I found this [blog post](http://theautomatic.net/2018/07/11/manipulate-files-r/) really helpful, it includes common tips related to manipulate files in r. 
Creating folders, we can use `file.create` function to accomplish 

<blockquote>
```{r, eval=FALSE}
file_name<-c(1:24)
file_location<-paste0(getwd(),'new',file_name)
file.create(file_location)
```
# Change Code content and move to desired location
The code I want to run is in `.sas` format. As I would like to change some code content like input dataset name, file location, etc, I would need to save the sas code as `.txt`, then use `readLines` to read in the txt file.  

In the txt file, I set up some keywords to help me to use `gsub` later on in order to modify the code slightly.  

For example, I changed `%let loc=/mnt/path/particular_file_loc;` to `%let loc=/mnt/path/DATA_LOCATION;`.   
In this way, I can use `gsub` to locate `DATA_LOCATION` and switch that to actual file directory.
Below is code

<blockquote>
```{r, eval=FALSE}
x <- readLines(sas_code_txt_format)
x[1]<-gsub('DATA_LOCATION',DATA_LOCATION, x[1])
x[7]<-gsub('Output_SASData_Name',Output_SASData_Name, x[7])
x[11]<-gsub('Output_SASCode_Name',Output_SASCode_Name, x[11])
#save txt to desired file location with desired code name 
write.table(x,file =Output_SAS_Code_Loc,row.names=F,col.names=F,quote = F)
```
# Run same UNIX command on different directory

For this particular task, I need to go to each sub-folders, and run same UNIX command `nohup sas code.sas &` within each sub-folder. 
Given we have 24 folders, I'd like to do it at once rather than type 24 `cd sub-folder` command in putty. By some research, I found this [stackoverflow post](https://stackoverflow.com/questions/7470165/how-to-go-to-each-directory-and-execute-a-command). 

To make ourselves life easier, make sure naming your folders in patterns, such as, I can make the folder always begin as `DDZR_Folder`, then I can use following UNIX code to run each code within each folder. 

>for d in ./DDZR_Folder*/  
>do  
>    ( cd "$d" && nohup sas sas_code.sas &  )  
>done  

Another helpful tips is run same command on different sas code within same directory  

> for filename in ./*.sas; do nohup sas $filename; done

One thing worth-noting is, it's better to do some experiment under a test folder before running UNIX command on the actual folder. 

# Convert xls to xlsx

We had some files are stored in xls, this [post](https://indranilgayen.wordpress.com/2016/05/11/batch-convert-xls-to-xlsx-using-r/) provides a way to convert all the xls to xlsx

>purrr::walk(all_24_folders_in_a_vector,convert_xls_to_xlsx) 


# Takeaway

This short blog post summarized how I automated a tedious task. It seems to know some UNIX command and regular expression would be really helpful. 




